### 2016-12-17 Chapters/theory.tex
Finishing chapter 3, sent to Hidde and Hans

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
\caption[Self-replicator model accounts for bacterial growth laws]{
\textbf{Self-replicator model accounts for bacterial growth laws.}\\
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

A self-replicator at steady state accumulates biomass according to $\texttt{Vol}(0)\, e^{\mu^*\, t}$, $t\in [0,\tau]$, when $\mu^*$ is the growth rate at steady state. 
The accumulation of biomass is obviously maximal when the growth rate is maximal ($\mu^* = \mu^*_{opt}$).
In dynamical conditions, the growth rate is not constant and biomass accumulation is described more generally by: 
\begin{equation*}
\frac{d\texttt{Vol}}{dt} = \mu(t) \, \texttt{Vol}.
\end{equation*}
In other words, when integrating over the time interval $[0,\tau]$:
\begin{equation}\label{eq:biomassdyn}
\ln \left(\frac{\texttt{Vol}(\tau)}{\texttt{Vol}(0)}\right) =  \int_{0}^{\tau} \mu(t) \, dt.
\end{equation}
Since the logarithm is an increasing function, maximizing the biomass produced over $[0,\tau]$ requires maximization of the right-hand side of the equation.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
\caption[Optimal control of the self-replicator during a nutrient upshift]{
{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
\caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{./Fig/Fig6.eps}
\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{./Fig/Fig7.eps}
\caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
{
{\bf Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}\newline
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig8.eps}
\caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
{
{\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

@@

\section{S1 Text -- Model derivation and analysis}
\manuallabel{S1_Text}{S1~Text}

\subsection{Model formulation}

@@

\subsection{Nondimensionalization of the system}

@@

\subsection{Steady-state growth of the self-replicator}
\label{si::growthrate}

@@

One can now easily derive the steady-state growth rate, denoted $\hat{\mu}^*$.
By substituting Eq.~\ref{eq:supp_growthrate_adim} into the first ODE of the system of Eq.~\ref{eq:supp_adim}, we find at steady state:
\[
\left(\frac{d\hat{p}}{d\hat{t}}\right)_{(\hat{p}^*, \hat{r}^*)} = 0 = (1-\alpha)\, E_M - (1+\hat{p}^*)\, \hat{\mu}^*,
\]
which by means of Eq.~\ref{eq:supp_steadystate} gives the following relation:
\begin{equation}
\label{eq:supp_growthrate_p}
\hat{\mu}^* = \frac{(1-\alpha)\, E_M}{1+\hat{p}^*} = \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
\end{equation}
Finally, we can transform this expression to obtain
\begin{equation}
\label{eq:supp_growthrate_final}
\hat{\mu}^* = \begin{cases}
\frac{(1-\alpha)\, E_M + \alpha - \sqrt{\left[(1-\alpha)\, E_M - \alpha\right]^2 + 4(1-\alpha)\, \alpha\, E_M\, K}}{2(1-K)} &\text{ for }K\neq 1,\\
\frac{\alpha\, (1-\alpha)\, E_M}{\alpha + (1-\alpha)\, E_M} &\text{ for }K = 1.
\end{cases}
\end{equation}
This function of $\alpha$ is plotted in Figure~\ref{fig:model_analysis}\textit{B} in the main text.

\subsection{Maximization of growth rate at steady state}
\label{si::optimal}

@@

\subsection{Analysis of the control strategies}
\label{si::control_strategies}

@@

\subsubsection{Nutrient-only strategy}

@@

\subsubsection{Precursor-only strategy}

@@

\subsubsection{On-off strategy}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
\caption
{
\textbf{Local stability of the on-off strategy.}
The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
The solid, black curve is the $\hat{p}$-nullcline.
The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
}
\label{fig:sliding}
\end{figure}

@@

\section{S2 Text -- Model parameters}
\manuallabel{S2_Text}{S2~Text}

@@

\subsection*{\Large \texorpdfstring{$e_M$}{eM}}

@@

\subsection*{\Large \texorpdfstring{$k_R$}{kR}}

@@

\subsection*{\Large \texorpdfstring{$K_R$}{KR}}

@@

\subsection*{\Large \texorpdfstring{$\beta$}{Beta}}

@@

\subsection*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}

@@

\section{S3 Text -- Solution of optimal control problem}
\manuallabel{S3_Text}{S3~Text}

\subsection{Statement of the problem}

@@

\subsection{Maximum Principle}
Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
Let $H(\hat{p},\hat{r},\lambda_p,\lambda_r,\lambda_0,\alpha)$ be the Hamiltonian of the system, defined by
\[
H(\cdot) \equiv \lambda_p\, E_M\, (1-\hat{r}) - \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}\left[\lambda_p (1+\hat{p}) +\lambda_r\, \hat{r} +\lambda_0\right] + \alpha \, \lambda_r \, \hat{r}\, \frac{\hat{p}}{K+\hat{p}}.
\]
Moreover, let $\alpha$ be an optimal control, and $\hat{x}(\cdot)=(\hat{p}(\cdot),\hat{r}(\cdot))$ the associated trajectory.
Then, there exists $\lambda_0 \leq 0$ and an absolutely continuous map $\lambda=(\lambda_p,\lambda_r):[0,+\infty) \rightarrow \mathbb{R}^2$
such that $(\lambda,\lambda_0)\neq0$, and
\begin{align}
\dot{\lambda}_p\, =-\frac{\partial H}{\partial \hat p}&= \hat{r} \, \frac{K}{(K+\hat{p})^2}\, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, ( \hat{r}-\alpha) + \lambda_0\right] + \hat{r}\, \frac{\hat{p}}{K+\hat{p}}\, \lambda_p, \label{eq:adjoint_p}\\
 \dot{\lambda}_r\, =-\frac{\partial H}{\partial \hat r}&= \lambda_p\, E_M +\frac{\hat{p}}{K+\hat{p}} \, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, (2 \hat{r}-\alpha) +\lambda_0\right] \label{eq:adjoint_r}.
\end{align}
The maximization condition is given by:
\begin{equation}{\label{eq:PMP}}
\begin{aligned}
&\alpha(\hat{t}) \in \arg\ max_{v \in [0,1]} H(\hat{x}(\hat{t}),\lambda(\hat{t}),\lambda_0,v), \\ &\mathrm{almost}\ \mathrm{everywhere}\ \mathrm{on}\ [0,+\infty).
\end{aligned}
\end{equation}

@@

\subsection{Characterization of singular arcs}

@@

\subsection{Analysis of the adjoint system}
Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}) may be useful to characterize the switches of extremal trajectories.

@@

\section[S4 Text -- Kinetic model of the ppGpp system]{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
\sectionmark{S4 Text -- Kinetic model of the ppGpp system}
\sectionmark{S4 Text -- Kinetic model of the ppGpp system}

\manuallabel{S4_Text}{S4~Text}

>>>>

\begin{figure}[tb]
\centering
\includegraphics[scale=0.9]{./Fig/Fig1.eps}
\caption{
\textbf{Self-replicator model of bacterial growth.}
External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\begin{figure}[tb]
\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
\caption{\textbf{Analysis of self-replicator model of bacterial growth.}
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
\caption{
\textbf{Self-replicator model accounts for bacterial growth laws.}
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

A self-replicator at steady state accumulates biomass according to \\ $\texttt{Vol}(0)\, e^{\mu^*\, t}$, $t\in [0,\tau]$, when $\mu^*$ is the growth rate at steady state. 
The accumulation of biomass is obviously maximal when the growth rate is maximal ($\mu^* = \mu^*_{opt}$).
In dynamical conditions, the growth rate is not constant and biomass accumulation is described more generally by: 
\begin{equation*}
\frac{d\texttt{Vol}}{dt} = \mu(t) \, \texttt{Vol}.
\end{equation*}
In other words, when integrating over the time interval $[0,\tau]$:
\begin{equation}\label{eq:biomassdyn}
\ln \left(\frac{\texttt{Vol}(\tau)}{\texttt{Vol}(0)}\right) =  \int_{0}^{\tau} \mu(t) \, dt.
\end{equation}
Since the logarithm is an increasing function, maximizing the biomass produced over $[0,\tau]$ requires maximization of the right-hand side of the equation.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
\caption{
\textbf{Optimal control of the self-replicator during a nutrient upshift.}
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
\caption{
\textbf{Alternative strategies for controlling the self-replicator of bacterial growth.}
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig6.eps}
\caption{
\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig7.eps}
\caption{
\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/Fig8.eps}
\caption{
\textbf{ppGpp regulation implements an on-off control strategy of resource allocation.}
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

@@

\section{Supporting information for Chapter~2}

\subsection{S1 Text -- Model derivation and analysis}
\manuallabel{S1_Text}{S1~Text}

\subsubsection{Model formulation}

@@

\subsubsection{Nondimensionalization of the system}

@@

\subsubsection{Steady-state growth of the self-replicator}
\label{si::growthrate}

@@

One can now easily derive the steady-state growth rate, denoted $\hat{\mu}^*$.
By substituting Eq.~\ref{eq:supp_growthrate_adim} into the first ODE of the system of Eq.~\ref{eq:supp_adim}, we find at steady state:
\[
\left(\frac{d\hat{p}}{d\hat{t}}\right)_{(\hat{p}^*, \hat{r}^*)} = 0 = (1-\alpha)\, E_M - (1+\hat{p}^*)\, \hat{\mu}^*,
\]
which by means of Eq.~\ref{eq:supp_steadystate} gives the following relation:
\begin{equation}
\label{eq:supp_growthrate_p}
\begin{split}
\hat{\mu}^* &= \frac{(1-\alpha)\, E_M}{1+\hat{p}^*}\\
 &= \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
\end{split}
\end{equation}
Finally, we can transform this expression to obtain
\begin{equation}
\label{eq:supp_growthrate_final}
\hat{\mu}^* = \begin{cases}
\frac{(1-\alpha)\, E_M + \alpha - \sqrt{\left[(1-\alpha)\, E_M - \alpha\right]^2 + 4(1-\alpha)\, \alpha\, E_M\, K}}{2(1-K)} &\text{ for }K\neq 1,\\
\frac{\alpha\, (1-\alpha)\, E_M}{\alpha + (1-\alpha)\, E_M} &\text{ for }K = 1.
\end{cases}
\end{equation}
This function of $\alpha$ is plotted in Figure~\ref{fig:model_analysis}\textit{B} in the main text.

\subsubsection{Maximization of growth rate at steady state}
\label{si::optimal}

@@

\subsubsection{Analysis of the control strategies}
\label{si::control_strategies}

@@

\paragraph{Nutrient-only strategy}

@@

\paragraph{Precursor-only strategy}

@@

\paragraph{On-off strategy}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
\caption
{
\textbf{Local stability of the on-off strategy.}
The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
The solid, black curve is the $\hat{p}$-nullcline.
The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
}
\label{fig:sliding}
\end{figure}

@@

\subsection{S2 Text -- Model parameters}
\manuallabel{S2_Text}{S2~Text}

@@

\subsubsection{\Large \texorpdfstring{$e_M$}{eM}}

@@

\subsubsection{\Large \texorpdfstring{$k_R$}{kR}}

@@

\subsubsection{\Large \texorpdfstring{$K_R$}{KR}}

@@

\subsubsection{\Large \texorpdfstring{$\beta$}{Beta}}

@@

\subsubsection{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}

@@

\subsection{S3 Text -- Solution of optimal control problem}
\manuallabel{S3_Text}{S3~Text}

\subsubsection{Statement of the problem}

@@

\subsubsection{Maximum Principle}

Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
Let $H(\hat{p},\hat{r},\lambda_p,\lambda_r,\lambda_0,\alpha)$ be the Hamiltonian of the system, defined by
\[
H(\cdot) \equiv \lambda_p\, E_M\, (1-\hat{r}) - \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}\left[\lambda_p (1+\hat{p}) +\lambda_r\, \hat{r} +\lambda_0\right] + \alpha \, \lambda_r \, \hat{r}\, \frac{\hat{p}}{K+\hat{p}}.
\]
Moreover, let $\alpha$ be an optimal control, and $\hat{x}(\cdot)=(\hat{p}(\cdot),\hat{r}(\cdot))$ the associated trajectory.
Then, there exists $\lambda_0 \leq 0$ and an absolutely continuous map $\lambda=(\lambda_p,\lambda_r):[0,+\infty) \rightarrow \mathbb{R}^2$
such that $(\lambda,\lambda_0)\neq0$, and
\begin{align}
\dot{\lambda}_p\, =-\frac{\partial H}{\partial \hat p}&= \hat{r} \, \frac{K}{(K+\hat{p})^2}\, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, ( \hat{r}-\alpha) + \lambda_0\right] + \hat{r}\, \frac{\hat{p}}{K+\hat{p}}\, \lambda_p, \label{eq:adjoint_p}\\
 \dot{\lambda}_r\, =-\frac{\partial H}{\partial \hat r}&= \lambda_p\, E_M +\frac{\hat{p}}{K+\hat{p}} \, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, (2 \hat{r}-\alpha) +\lambda_0\right] \label{eq:adjoint_r}.
\end{align}
The maximization condition is given by:
\begin{equation}{\label{eq:PMP}}
\begin{aligned}
&\alpha(\hat{t}) \in \arg\ max_{v \in [0,1]} H(\hat{x}(\hat{t}),\lambda(\hat{t}),\lambda_0,v), \\ &\mathrm{almost}\ \mathrm{everywhere}\ \mathrm{on}\ [0,+\infty).
\end{aligned}
\end{equation}

@@

\subsubsection{Characterization of singular arcs}

@@

\subsubsection{Analysis of the adjoint system}

Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}) may be useful to characterize the switches of extremal trajectories.

@@

\subsection[S4 Text -- Kinetic model of the ppGpp system]{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
\sectionmark{Kinetic model of the ppGpp system}
\manuallabel{S4_Text}{S4~Text}
